---
title: "Examining the Effects of "
author: "Taeil Goh, Greg Tozzi & Max Ziff"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: github_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    output_file = 'README.md',
    inputFile, encoding = encoding,
    output_format = c('github_document')) 
    })
---

```{r setup, include=FALSE}

library(data.table)
library(foreign)
library(lmtest)
library(sandwich)
library(stargazer)
library(data.table)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

`place holder`

## Introduction and Context 

#### The Children’s Science Center

The Children’s Science Center (“the Center”) is a non-profit organization headquartered in Fairfax, Virginia.  Founded in TODO, the Center’s principal objective is to build the Northern Virginia region’s first interactive science center at a greenfield location on donated land in Reston, Virginia.  The capital campaign to raise fund to build this facility is the Center’s primary line of effort.  A recent partnership with the Commonwealth of Virginia’s 

Recognizing the long-term nature of the capital campaign and the need to build interest and advocacy, the Center launched a series of intermediate efforts beginning in `TODO.`  The Center’s first outreach effort was the Museum without Walls, a program that delivered science-based programs to schools around the region.  In `TODO`, the Center launched the Lab, a test facility located in a shopping mall sited 12 miles outside of the Washington, DC Beltway.  The Lab hosted approximately `TODO` guests per year until it was temporarily shut down in response to the COVID-19 pandemic.

The Center funds operating expenses with Lab admissions, grants, and philanthropic giving.  The Center’s Development Department is responsible for developing and managing the latter two revenue streams.

##### The Fundraising Surge

The Center conducts a fundraising surge toward the end of the calendar year, consistent with the charitable giving cycle in the United States [`TODO`: add reference].

The Center expected to engage the majority of its stakeholders by email.  The Center viewed email marketing as repeated iterations of a three step process.  The Development Director expected that several rounds of engagement would be necessary to achieve a conversion in the form of a charitable gift.  Within each individual round, the path to conversion proceeded from receipt to opening to click-through using an embedded link or button.

The Center introduced two major changes to its process ahead of the 2020 fundraising surge.  In previous year-end campaigns, the Center had used Constant Contact to send branded emails to stakeholders.  In the 2020 campaign, the Center chose to switch to Mail Chimp’s free tier.  The Center also engaged a consulting firm that applies machine learning to the nonprofit space to predict top prospects from among the Center’s contact list and to predict affinities to aspects of the Center’s mission.  The consulting firm’s model requires a physical address as an entering argument, and it was, therefore, only applied to that subset of the Center’s contact list for which physical addresses were entered.  Based on the consulting firm’s output, the Center chose to engage a subset of their stakeholder list with physical mail and to wall this cohort off from subsequent email engagement.

#### Research Questions

The Center's development staff was most strongly interested in conducting experiments to maximize the opening rates of the Center's fundraising emails.  The Development Director and her staff were keenly aware that email opening was the critical first step in achieving conversion.  The follow-on actions---click though and conversion---were areas of secondary interest for this study.  With this in mind, we sought to answer two specific questions posed by the Center.

1. Is there a difference in email opening or click-through caused by using the Executive Director’s name and title in the from line of the solicitation email versus using the Board Chair’s name and title?

2. Is there a difference in email opening or click-through behavior caused by the choice of using one of two potential subject lines?  The two subject lines considered were, “You can be a Catalyst for STEM Learning 💡," and , "Invest in the Power of STEM Learning 💡."


## Research Proposal 

#### Research Hypotheses


We claim expertise neither in non-profit email fundraising nor in the preferences of the Center's stakeholders.  The Center's development team believed that there would be a significant difference in the opening rates associated with the two email from lines considered, though the staff did not have a sense of which would result in significantly higher opening rates.  Underpinning the staff's expectation that we would find a significant difference was their belief that the Center's stakeholders had a meaningful sense of the Center's leadership and governance structure and would react differently when presented with emails from either the Executive Director or the Board Chair.

Both of the subject lines considered were written by the Center's development staff.  The staff believed that subject lines could result in significantly different opening rates but did not have a going-in belief of which subject line would cause a stronger response.

#### Treatment in Details 
`see page 428`

The experiment was structured as two concurrant A/B tests conducted on a single subject cohort.  With the cohort randomly assigned to four groups, and with the from and subject lines denoted $F_a$, $F_b$, $S_a$, and $S_b$, we assigned treatment combinations across balanced groups:

- Group 1 - $F_a$ and $S_a$
- Group 2 - $F_a$ and $S_b$
- Group 3 - $F_b$ and $S_a$
- Group 4 - $F_b$ and $S_b$

This design allows us to explore heterogeneous treatment effects.

Each group would received a tailored email containing its assigned treatments sent through Mail Chimp.  We intended to track opening and click through using Mail Chimp's out-of-the-box analytics.

#### Power calculation 

`place holder`

```{r}

```


#### Enrollment Process & Criteria for Subjects
- The criteria by which subjects were included
- explain how subjects were randomly assigned to experimental groups
- TG: need explanation of 2000 selection was randomized and no selection bias 

The Center provided an anonymized list of TODO stakeholders extracted from its contact database.  The individuals represented the Center's complete contact list less a cohort of approximately 2,000 individuals identified by the Center for a special traditional mail campaign and less a relatively small number of individuals walled off from fundraising activities due to age, relationship to board members, and other similar factors.  The Center provided the output of the consulting firm's model for individuals with mailing address on file.  This output consisted of buckted predictions of affinity for individuals to certain causes related to the Center's mission---educational causes, children's causes, cultural causes, and scientific causes.

We agreed to target a random sample of 1,980 individuals for this experiment.  The intent was to remain within the bounds of the Mail Chimp free tier while leaving a small margin for the Center's staff to send test emails to individuals outside of the experiment including key members of the Center's staff and our group. The Center delivered one of four solicitation emails via Mail Chimp's web interface.


#### Summary of Experimental Design

#### Validation of Randomization Procedure
- check the soundness of the randomization procedure 
- covariate balance (see chapter 4)

Balance checks using Boodle.ai data

#### Outcome Measures

- Open and click-through rates.
- mail chimp 

#### Plan to analyze the data

- how you plan to analyze the data 

#### Accounting for Non-Compliance



## Research Report

#### Flow Diagram 

`optional`

#### Experiment Results 

- Attrition 
- additional outcome (e.g., unsubscription rate)
- Mediation analysis: why more open rate? 
- description of experimental results 

We are confident that the method we employed in this case produced evidence that the difference in opening rates was caused by the difference in subject lines.  You can reasonably expect that this result would generalize to the remainder of the candidate email recipients during the current year-end fund drive.  We suggest caution when drawing broader conclusions, however for the following reasons:
1. Responsiveness to the two subject lines may have been affected by factors that we could not control for, including the current public health and political situations.
2. We drew our subjects from a filtered list of potential recipients.  Adding individuals to the population of recipients may invalidate the study’s results.


```{r helper function}

calculate_rse_ci <- function(lm.mod, dt, alpha = 0.05){
    lm.mod$se.ci_ <- confint(lm.mod, level = 1 - alpha)
    lm.mod$vcovHC_ <- vcovHC(lm.mod)
    lm.mod$rse_ <- sqrt(diag(lm.mod$vcovHC_))
    lm.mod$rse.ci_ <- confint(coeftest(lm.mod, vcov. = lm.mod$vcovHC_), level = 1 - alpha)
    
    return(lm.mod)
}

```

```{r cars}
#d <- fread("./data/20201025_pilot_data.csv")
d <- fread("../data/raw/20201029_results_with_covariates")

d[treatment_assigned == 1 | treatment_assigned == 2, subject := 1]
d[treatment_assigned == 3 | treatment_assigned == 4, subject := 0]
d[treatment_assigned == 1 | treatment_assigned == 3, sender := 1]
d[treatment_assigned == 2 | treatment_assigned == 4, sender := 0]

#summary(d)
#head(d)
#names(d)
#d = na.omit(d)
```


```{r warning=FALSE}
m1 = d[ , lm(open ~ subject)]
m2 = d[ , lm(open ~ sender)]
m3 = d[ , lm(open ~ subject + sender + subject*sender)]
m1 = calculate_rse_ci(m1, d)
m2 = calculate_rse_ci(m2, d)
m3 = calculate_rse_ci(m3, d)

stargazer(m1, m2, m3, se = c(list(m1$rse_), list(m2$rse_), list(m3$rse_)), type="text")
    
```
```{r}
m3$rse.ci_
```



