---
title: "Increase Donation with Email Campaign"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: github_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    output_file = 'README.md',
    inputFile, encoding = encoding,
    output_format = c('github_document')) 
    })
    
---

## Research Questions
We sought to answer two questions posed by the Center.  Both questions were related to emails sent by the Center soliciting donations.

1. Is there a difference in email opening or click-through caused by the choice of using the Executive Director’s name and title in the from line versus using the Board Chair’s name and title?  While we did observe a difference in opening rates favoring emails from Jill McNabb, the difference was not significant.

2. Is there a difference in email opening or click-through behavior caused by the choice of using one of two potential subject lines?  The use of the subject line, “You can be a Catalyst for STEM Learning 💡,” caused significantly higher opening rates.

## Review of the Methodology
The Center provided our team with a list of approximately 12,000 potential subjects for this study.  This list was compiled from the Center’s Altru database but specifically excluded individuals who were previously selected by the Center to receive a solicitation by traditional mail.  We randomly selected 1,980 subjects to receive an email soliciting a donation.  These individuals were evenly split into four groups of 495 with each group being assigned a unique combination of from and subject lines.  The 68% of the population that had mailing addresses on file with the Center also had metadata provided by Boodle.ai related to their predicted affinity for various causes.  We used this metadata to check the balance of our random assignment


## Pre Experiment 
`something related to preparing the experiment`

#### Power calculation 

`place holder`

```{r}

```

#### Covariate Balance Check (Pre-experiment)
`place holder`

## Post Experiment 

#### Measurement 


```{r setup, include=FALSE}

library(data.table)
library(foreign)
library(lmtest)
library(sandwich)
library(stargazer)
library(data.table)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

```{r helper function}

calculate_rse_ci <- function(lm.mod, dt, alpha = 0.05){
    lm.mod$se.ci_ <- confint(lm.mod, level = 1 - alpha)
    lm.mod$vcovHC_ <- vcovHC(lm.mod)
    lm.mod$rse_ <- sqrt(diag(lm.mod$vcovHC_))
    lm.mod$rse.ci_ <- confint(coeftest(lm.mod, vcov. = lm.mod$vcovHC_), level = 1 - alpha)
    
    return(lm.mod)
}

```


## Result Analysis

- Subject: 6 percentage point [0.3pp ~ 11pp]
- Sender: no significant difference

```{r cars}
#d <- fread("./data/20201025_pilot_data.csv")
d <- fread("../data/raw/20201029_results_with_covariates")

d[treatment_assigned == 1 | treatment_assigned == 2, subject := 1]
d[treatment_assigned == 3 | treatment_assigned == 4, subject := 0]
d[treatment_assigned == 1 | treatment_assigned == 3, sender := 1]
d[treatment_assigned == 2 | treatment_assigned == 4, sender := 0]

#summary(d)
#head(d)
#names(d)
#d = na.omit(d)
```
Lets quickly check the open treatment effect. 

```{r warning=FALSE}
m1 = d[ , lm(open ~ subject)]
m2 = d[ , lm(open ~ sender)]
m3 = d[ , lm(open ~ subject + sender + subject*sender)]
m1 = calculate_rse_ci(m1, d)
m2 = calculate_rse_ci(m2, d)
m3 = calculate_rse_ci(m3, d)

stargazer(m1, m2, m3, se = c(list(m1$rse_), list(m2$rse_), list(m3$rse_)), type="text")
    
```
```{r}
m3$rse.ci_
```


#### Notes on Applying the Results
We are confident that the method we employed in this case produced evidence that the difference in opening rates was caused by the difference in subject lines.  You can reasonably expect that this result would generalize to the remainder of the candidate email recipients during the current year-end fund drive.  We suggest caution when drawing broader conclusions, however for the following reasons:
1. Responsiveness to the two subject lines may have been affected by factors that we could not control for, including the current public health and political situations.
2. We drew our subjects from a filtered list of potential recipients.  Adding individuals to the population of recipients may invalidate the study’s results.

#### Potential Additional Topics
- Attrition 
- additional outcome (e.g., unsubscription rate)
- Mediation analysis: why more open rate? 
